---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# caladaptR <img src="man/figures/caladaptr-beta_logo.svg" align="right" width="240" />

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
<!-- badges: end -->

`caladaptr` is an API client that makes it easier to work with data from [Cal-Adapt.org](https://cal-adapt.org/) in R.

'Beta' status means the package:

1) is still under development  
2) is being updated more often than usual  
3) has a slightly higher probability than average that updates will *not* be backward compatible  
4) is actively seeking user feedback and input (interested? join the [caladaptR betaR](#caladaptr-betar-club) club!)

## Installation

`caladaptr` is hosted on [GitHub](https://github.com/UCANR-IGIS/caladaptr){target="_blank" rel="noopener"}. To install it, you need to have [RTools](https://cran.r-project.org/bin/windows/Rtools/){target="_blank" rel="noopener"} and the `devtools` package installed:

```{r install_caladaptr, eval = FALSE}
library(devtools)
devtools::install_github("ucanr-igis/caladaptr")
```

## Example 1: Get Annual Projected Temperature at a Point Location

1) Load the package:

```{r load_package}
library(caladaptr)
```

2) Next, create an **API request object** for a point location. This doesn't actually fetch any data, it puts together the pieces  pieces of an API request. There are a number of '*constructor*' functions you can mix and match to create an API request object, for example:

```{r cap1_create, cache = TRUE}
cap1 <- ca_loc_pt(coords = c(-121.4687, 38.5938)) %>%  ## specify a location
  ca_gcm(gcms[1:4]) %>%                                ## select GCM(s)
  ca_scenario(scenarios[1:2]) %>%                      ## select emission scenarios(s)
  ca_period("year") %>%                                ## select a precooked temporal aggregation period
  ca_years(start = 2040, end = 2060) %>%               ## select start and end dates
  ca_cvar(c("tasmax", "tasmin"))                       ## select climate variables

cap1
```

To help you pass arguments for the various constructor functions, `caladpatr` provides the following constants:

```{r constants, cache = TRUE}
climvars

gcms

## Note the first four GCMs are the 'priority' models recommended under 
## California's 4th Climate Change Assessment.

scenarios

periods

```

**Note**: Cal-Adapt has data for many but by no means all permutations of the above constants.

3) Next, feed your API call into a function that **fetches data**, such as `ca_getvals()`. 

```{r cap1_getvals, cache = TRUE}
cap1_vals_df <- ca_getvals(cap1, quiet = TRUE) %>% ca_vals2tbl()
dim(cap1_vals_df)
head(cap1_vals_df)
```

**Note #1**. values are returned with units, thanks to the [units](https://cran.r-project.org/package=units) package. No more guessing what the units are, or having to lookup unit conversion constants!

**Note #2**. See how the results from `ca_getvals` get fed into `ca_vals2tbl` to become a tibble? See `?ca_vals2tbl` for more options to specify which column(s) to return.

## Example #2: Same query as #1, but for a Congressional District

Cal-Adapt supports ~10 'preset' Areas of Interest. `caladaptr` refers to these as "AOI Presets', and you can use them to specify area(s) of interest in your API requests:

```{r}
aoipreset_types
```

Next, we'll modify the first API call by simply "swapping out" the point location for a Congressional District location, using the `ca_loc_aoipreset()` function. For this example we'll use California's 20th and 24th [Congressional Districts](https://en.wikipedia.org/wiki/California%27s_congressional_districts) (Carmel Valley and Santa Barbara). 

```{r cap2_define, cache=TRUE}
cap2 <- cap1 %>% ca_loc_aoipreset(type = "cdistricts",
                                  idfld = "geoid",
                                  idval = c("0620", "0624"),
                                  stat = "mean")

cap2
```

Note when you query a polygon, you have to also specify a summary statistic that will be used to reduce the various pixels that fall within the polygon to one number. Your choices are `mean`, `max`, `median`, `min`, and `sum.` (If you want to retain the individual pixel values, use `ca_getrst()`.)

```{r ca2_getvals, cache=TRUE}
cap2_vals_df <- ca_getvals(cap2, quiet = TRUE) %>% ca_vals2tbl()

dim(cap2_vals_df)
head(cap2_vals_df)
```

# caladaptR betaR club

The **caladaptR betaR** club is **now accepting members!** If you would like to be part of this elite cadre of `caladaptr` early adopters, you should join our beta testing program. Benefits of joining the **caladaptR betaR** club include:

- early notification of **package updates**  
- 24/7 **priority support**  
- **exclusive invitations** to `caladaptr` tutorials and user feedback sessions
- an **email forum** to bounce ideas around with other caladaptR users and the package developer

To join the **caladaptR betaR** club, please subscribe yourself to the [caladaptR email list](https://lists.ucdavis.edu/sympa/subscribe/caladaptr).

# Coming soon...

 - How to query a `sf` object  
 - More examples of reshaping data  
 - Retrieving rasters  

