---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# caladaptr

<!-- badges: start -->
<!-- badges: end -->

`caladaptr` is an API client that makes it easier to work with data from Cal-Adapt.org in R.

## Installation

Currently, caladaptr is hosted on [GitHub](https://github.com/UCANR-IGIS/caladaptr){target="_blank" rel="noopener"} in a private repository. To install it, you need to i) be added to the repo as a collaborator (contact Andy Lyons or Shane Feirer), and ii) generate a [personal access token](https://github.com/settings/tokens){target="_blank" rel="noopener"} for yourself. A personal access token is a bit like a password, which you feed into the 'install_github' as shown below.


```
library(devtools)
install_github("UCANR-IGIS/caladaptr", auth_token = "insert-your-token-here")
```

## Usage

```{r example}
library(caladaptr)
## basic example code
```


