---
title: "Intro to caladaptR"
output: html_notebook
---

# About this R Notebook

R Notebooks are a 'flavor' of R markdown that combine plain text with R commands in code chunks. You can (and should!) edit them, and save it every few minutes like anything else!

If you're in RStudio, go ahead and *minimize the console window*. You won't need it. When you run the R commands the *output appears below the code chunk*.

Keyboard shortcuts:
 - run the current line of R: *ctrl + enter*  
 - run everything in the current code chunk: *ctrl + shift + enter*  
 - insert a new code chunk: *ctrl + alt + i*  

# Setup

## Install any missing packages

```{r}
pkgs_req <- c("remotes", "ggplot2", "dplyr", "tmap", "conflicted", "leaflet", "tidyr")
pkgs_missing <- pkgs_req[!(pkgs_req %in% installed.packages()[,"Package"])]
if (length(pkgs_missing)) install.packages(pkgs_missing, dependencies=TRUE)
```
## Install caladaptr

```{r}
remotes::install_github("ucanr-igis/caladaptr")
```

## Load all packages

Load other required libraries we'll use in the sample code below:

```{r}
library(caladaptr)
library(units)
library(ggplot2)
library(dplyr)
library(conflicted)
library(tidyr)
library(tmap)
library(sf)
```

## Set preferences for common function names 

Set conflict preferences to prevent errors with common `dplyr` functions:

```{r}
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("count", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
```


