---
title: "Introduction"
output:
  html_notebook:
    df_print: paged
    toc: yes
    toc_float: yes
---

```{css echo = FALSE}
h1 {
  font-weight: bold;
  font-size: 24px;
  color: darkolivegreen;
  border-top: 3px solid dimgrey;
  margin-top: 1em;
  padding-top: 0.5em;
}
h1.title {
  color: black;
  border: none;
}
h2 {
  font-weight: bold;
  font-size: 22px;
  color: dimgray;
}

h3 {
  font-weight: bold;
  font-size: 18px;
  color: black;
}

```

<p style="text-align:right;"><img src="https://ucanr-igis.github.io/caladaptr/reference/figures/caladaptr-beta_logo.svg" width="240" /></p>

# About this R Notebook

R Notebooks are a 'flavor' of R markdown that combine plain text and R commands in code chunks. You can download the Rmd file from the 'code' button at the top of the page, then open it in RStudio. You run code chunks in the document line-by-line, and the output appears immediately below the code chunk. When you save the Rmd file, it automatically creates a HTML file showing results from all the code that has been run so far.

If you're in RStudio, you can *minimize the console window* (and probably close the right-hand panes as well). You won't need it,  because when you run R commands in a R Notebook the *output appears below the code chunk* (not the console). This takes some getting used to.

Keyboard shortcuts:  
- run the current line of R: *ctrl + enter*  
- run everything in the current code chunk: *ctrl + shift + enter*  
- insert a new code chunk: *ctrl + alt + i*  

# Setup

Install `caladaptr`:

```{r}
if (!require(devtools)) {install.packages("devtools"); library(devtools)}
# 
# devtools::install_github("ucanr-igis/caladaptr")
library(caladaptr)
## Make sure you have version 0.4.1 caladaptr or later. 
packageVersion("caladaptr")
```

Now we can load the other packages we'll be using below:

```{r library_all, message = FALSE}
pkgs_req <- c("ggplot2", "dplyr", "tidyr", "lubridate", "conflicted")
pkgs_missing <- pkgs_req[!(pkgs_req %in% installed.packages()[,"Package"])]
if (length(pkgs_missing)) install.packages(pkgs_missing, dependencies=TRUE)

library(units)
library(ggplot2)
library(dplyr)
library(conflicted)
library(tidyr)
library(tmap)
library(lubridate)
library(sf)
```

The last setup task is to define your preferences when you use an ambiguous function name (i.e., a function that exists in more than one package). This is particularly needed with a few common generic functions from `dplyr`:

```{r set_conflicts}
conflict_prefer("filter", "dplyr", quiet = TRUE)
conflict_prefer("count", "dplyr", quiet = TRUE)
conflict_prefer("select", "dplyr", quiet = TRUE)
```

# Example #1: Get Projected Temperature for a Point

**Goal**: Make a **time series plot** of projected maximum annual temperature for a **single point**, using the **four recommended GCMs** for California under RCP 4.5.

## 1\. Create the API Request 

The first step in getting climate variables back is to create a Cal-Adapt API request object. This involves stringing together a series of functions that specify the pieces of the request. The following constants can help you specify pieces of the request:

```{r}
cap1 <- ca_loc_pt(coords = c(-121.4687, 38.5938)) %>%
  ca_gcm(gcms[1:4]) %>%                                 
  ca_scenario(c("rcp45", "rcp85")) %>%
  ca_period("year") %>%
  ca_years(start = 2030, end = 2080) %>%
  ca_cvar(c("tasmax"))
```

Entering the name of an API request object at the console will display its components:

```{r}
cap1
```


*Tip*: You don't have to memorize a bunch of keywords. caladaptr provides several constants that contain the values you can pass to API construction functions, including `gcms`, `scenarios`, `cvars`, and `periods`.

```{r view_constants}
## Available global change models
## Note, the first 4 GCMs are the recommended priority models for California
gcms

## Available emissions scenarios
scenarios

## Available climate variables
cvars

## Available temporal aggregation periods
periods
```

**Note**: Cal-Adapt has data for many but by no means all combinations of the above constants.

You can use these values as arguments to build up the API request:



To verify the location in an API request, you can plot it. (Note we still haven't fetched any climate data yet, this just shows you the location the request will ask for.)

```{r plot_cap1}
plot(cap1)
```

## 2\. Fetch Data

Now it's time to fetch data with `ca_getvals_tbl()`. The object returned by `ca_getvals_tbl()` is tibble (data frame):  

```{r cap1_fetch, cache = TRUE}
cap1_tbl <- cap1 %>% 
  ca_getvals_tbl(quiet = TRUE)

cap1_tbl
```

## 3\. Munge the Results

To produce the desired time series plot, we need to i) pull out just values for RCP 4.5, and ii) convert degrees to &#176;F. For the unit conversion, we can use the handy `set_units` function from the `units` package.

```{r cap1_munge}
cap1_rcp45_tbl <- cap1_tbl %>%
  filter(scenario == "rcp45") %>%
  mutate(temp_f = set_units(val, degF))

cap1_rcp45_tbl
```

## Plot the Time Series

Plot these with ggplot:

```{r ggplot_cap1, cache = TRUE}
ggplot(data = cap1_rcp45_tbl, aes(x = as.Date(dt), y = as.numeric(temp_f), group = gcm)) +
  geom_line(aes(color=gcm)) +
  labs(title = "Average Annual Maximum Temperature for RCP4.5", x = "year", y = "temp (F)")
```

## YOUR TURN

Modify the above to create a similar plot for RCP 8.5.

```{r}
## Plot of RCP 8.5 goes here

```

# Example #2: Retrieve County Data 

*Goal*: produce a **time series plot** showing the **difference between RCP85 and RCP45** on maximum annual temperature for a **single county**.

The Cal-Adapt API has a number of 'preset' areas-of-interest (also called boundary layers) that you can use cookie-cutter style when retrieving climate data. The advantage of using an AOI Preset is that you don't need to import a GIS layer to query according to these common features. You just need to know the name or ID number of the feature(s) you're interested in.

The following AOI Presets are available:

```{r preset_types, cache = FALSE}
aoipreset_types
```

## 1\. Find the FIPS code for your county of interest

To use an AOI Preset, you need to specify which feature(s) you're interested in by providing the value(s) for one of the id fields. The specific columns available for identifying features vary according to preset. You can view the id columns and their values for an AOI Preset using the built-in `aoipreset_idval` constant. For example the counties layer allows you to specify a county by name, fips code, or id. Remember that everything in R is case-sensitive!

```{r idval_counties, cache = TRUE}
aoipreset_idval$counties
```

You can find county fips codes on Google - just be sure to use the 6-character version which includes the state. Alternately, you can plot the county preset layer and click on your county of interest: 

```{r plot_counties, message = FALSE}
counties_sf <- ca_aoipreset_geom("counties")
tmap_mode("view")
tm_shape(counties_sf) + tm_polygons()
```

For this example, we'll look at **Kings County** (FIPS = `06031`).

## 2\. Create the API Request

Let's create an API request for Kings County. Note below the inclusion of `ca_options()` to specify how we want to aggregate the pixels that fall within the country. This is required whenever you query polygon features.

```{r cap2_define, cache = TRUE}
cap2 <- ca_loc_aoipreset(type="counties", idfld = "fips", idval = "06031") %>%
  ca_gcm(gcms[1:4]) %>%
  ca_scenario(c("rcp45", "rcp85")) %>%
  ca_period("year") %>%
  ca_years(start = 2030, end = 2080) %>%
  ca_cvar(c("tasmin")) %>%
  ca_options(spatial_ag = "max")

cap2
```

As before, we can plot the API request to double-check we got the right location: 

```{r plot_cap2, cache = TRUE}
plot(cap2)
```

## 3\. Fetch Data

Fetch data with `ca_getvals_tbl()`:

```{r cap2_fetch, cache = TRUE}
cap2_tbl <- cap2 %>% 
  ca_getvals_tbl(quiet = TRUE) 

cap2_tbl
```

## 4\. Munge Results

To compute the difference between RCP 8.5 and RCP 4.5, we need to split them into separate columns. This is an example of going from a 'long' format to a 'wide' format. Fortunately, the tidyr package has a function called `pivot_wider` that can do this in single command. While we're at it we'll convert the temp to degrees Fahrenheit:

```{r cap2_wider, cache = TRUE}
rcp85_minus_rcp45_tbl <- cap2_tbl %>% 
  mutate(temp_f = set_units(val, degF)) %>% 
  select(fips, gcm, scenario, dt, temp_f) %>% 
  pivot_wider(names_from = scenario, values_from = temp_f)

head(rcp85_minus_rcp45_tbl)
```

## 5\. Plot 

Now we're ready to make the plot:

```{r cap2_ggplot, cache = TRUE}
ggplot(data = rcp85_minus_rcp45_tbl, aes(x = as.Date(dt), y = as.numeric(rcp85 - rcp45), group = gcm)) +
  geom_line(aes(color=gcm)) +
  labs(title = "Difference between RCP8.5 and RCP4.5 in the Maximum Annual Minimum \nTemperature for Kings County", x = "year", y = "temp (F)")
```

# Example #3: Analyze Extreme Heat in a Census Tract

*Goal*: Compare extreme heat occurrence under RCP 4.5 and RCP 8.5 for a single census tract over a 20-year time span. We'll measure 'extreme heat' in two ways, i) the proportion of total days when the max temp exceeds a certain threshold, and ii) the number of 'heat spells' defined as a period of *n* or more consecutive days when the temperature exceeds a threshold.

## 1\. Find the GEOID of a Census Tract in Kings County

Census tracts have 10-digit GEOID numbers rather than 6-digit FIPS codes ([more info](https://www.census.gov/programs-surveys/geography/guidance/geo-identifiers.html){target="_blank" rel="noopener"}). To find the GEOID for a census tract in Kings County, we could:

1) Ask Google.  
2) Plot all the census tracts (as we did above with counties), and click on the one we're interested in.  
3) Look it up using the `tidycensus` package.  
4) Take advantage of the fact that the GEOID of census tracts in Kings County all start with its FIPS code (6031).  

We'll use method #4, and grab the census tracts GEOIDs from `aoipreset_idval`:

```{r}
tracts_kings <- grep(pattern = "^6031", aoipreset_idval$censustracts$tract, value = TRUE)
tracts_kings
```

To plot just these tracts we can feed these GEOID values into `filter()`:

```{r plot_kings, message = FALSE}
census_tracts_sf <- ca_aoipreset_geom("censustracts")
tmap_mode("view")
tm_shape(census_tracts_sf %>% filter(tract %in% tracts_kings)) + tm_polygons()
```

For the rest of this exercise we'll use **6031001701**.

## 2\. Construct the API Request

The following will give us 20 years of daily maximum temperature for our census tract. To keep it simple we'll just use one GCM.

```{r cap3_define, cache = TRUE}
cap3 <- ca_loc_aoipreset(type="censustracts", idfld = "tract", idval = 6031001701) %>%
  ca_gcm("MIROC5") %>%
  ca_scenario(c("rcp45", "rcp85")) %>%
  ca_period("day") %>%
  ca_years(start = "2070-01-01", end = "2089-12-31") %>%
  ca_cvar("tasmax") %>% 
  ca_options(spatial_ag = "max")

cap3
```

```{r}
plot(cap3)
```

## 3\. Fetch Data

Now we're ready to fetch data:

```{r cap3_fetch, cache = TRUE}
cap3_tbl <- cap3 %>% 
  ca_getvals_tbl(quiet = TRUE)

dim(cap3_tbl)
head(cap3_tbl)
```

## 4\. Munge the Results

The only 'munging' we need to do is to add a column for the temperature in Farenheit.

```{r}
cap3_degf_tbl <- cap3_tbl %>% mutate(temp_f = set_units(val, degF))

head(cap3_degf_tbl)
```

## 5\. Compare the Distribution of Maximum Daily Temp

Plot a histogram of maximum temperature over this 20-year period for each emission scenario:

```{r cap3_vals_hist1, cache=TRUE}
hist(cap3_degf_tbl %>%  filter(scenario == "rcp45") %>% pull(temp_f),
     main = "RCP 4.5: Maximum Daily Temp, 2070-2089",
     xlab = "temp (F)") 

hist(cap3_degf_tbl %>%  filter(scenario == "rcp85") %>% pull(temp_f),
     main = "RCP 8.5: Maximum Daily Temp, 2070-2089",
     xlab = "temp (F)") 

```

## 6\. Count the Number of Days over 110&#176;F Projected Under each RCP

```{r}
cap3_degf_tbl %>% 
  mutate(really_hot_tf = temp_f >= set_units(110, degF)) %>%
  group_by(scenario, really_hot_tf) %>% 
  count() %>% 
  pivot_wider(names_from = scenario, values_from = n)
```

## 7\. Count the Number of Heat Spells Projected Under each RCP

A simple definition of 'heat spell' is *consecutive hot days*. Let's count Count the number of heat spells consisting of **4 or more consecutive days** of >110&#176;F.

The first step is to pull out the values for just one RCP, sort them by date, add a logical column if the maximum temp that day exceeded our treshhold. We'll then feed those values into `rle()`, which breaks up the time series into 'runs' of TRUE and FALSE.

```{r rcp45_runs, cache = TRUE}
## Use rle() to chop up a time series of TRUE of FALSE into 'runs'
rcp45_heat_runs <- cap3_degf_tbl %>% 
  filter(scenario == "rcp45") %>% 
  arrange(dt) %>% 
  mutate(really_hot_tf = temp_f >= set_units(110, degF)) %>% 
  pull(really_hot_tf) %>% 
  rle()

rcp45_heat_runs
```

Next, we find the runs where two conditions are met: run length is >= 4 days, and the run value = TRUE (meaning its a run of really hot days).

```{r}
sum(rcp45_heat_runs$values & rcp45_heat_runs$lengths >= 4)
```

We can do the same for RCP 8.5:

```{r rcp85_runs, cache = TRUE}
rcp85_heat_runs <- cap3_degf_tbl %>% 
  filter(scenario == "rcp85") %>% 
  arrange(dt) %>% 
  mutate(really_hot_tf = temp_f >= set_units(110, degF)) %>% 
  pull(really_hot_tf) %>% 
  rle()

sum(rcp85_heat_runs$values & rcp85_heat_runs$lengths >= 4)
```

## Conclusion

Under RCP 4.5, the MIROC5 model predicts this census tract will experience **36 extreme heat spells** (defined as 4 or more days where the temperature hit 110&#176;F) during this 20 year period. In comparison, MIROC5 predicts **45 extreme heat spells** under RCP 8.5.


## YOUR TURN

How many heat spells will there be under each RCP if we define a heat spell to be at least **7 days** over **100 &#176;F**?

```{r}
## Your answer goes here:

```

Plot the distribution of the number of consecutive days over 100 &#176;F (i.e., heat spell duration).

```{r}
## Your answer goes here:

```


